# A vocabulary for characterizing interactive graphics

This tutorial presents several R packages for making interactive
graphics. To begin, make sure to install the required packages:

```{r}
source("http://tdhock.github.io/interactive-tutorial/packages.R")
```

## Definition of interactive graphics

The R command line is interactive, so aren't all R graphics
interactive? Yes, if you have access to the data and R source
code. No, if you only have access to the graphic itself.

This tutorial uses the following definition: in an **interactive** or
**dynamic** graphic, the plot can be updated without changing the R
code. In a **non-interactive** or **static** graphic, the R code must
be changed to update the plot.

For example, the following is a static graphic

```{r}
lungDeaths <- cbind(mdeaths, fdeaths)
matplot(lungDeaths, type="l")
```

because it can be zoomed only by changing the R code:

```{r}
matplot(lungDeaths, type="l", xlim=c(20, 40))
```

In contrast, the following interactive graphic can be zoomed by
dragging the mouse on the plot (without changing the R code).

```{r}
library(dygraphs)
dygraph(lungDeaths)
```

We will use the terms **Actions** and **Effects** to categorize
different kinds of interactive graphics. In the graphic above, we see
several examples of **direct manipulation actions** which produce the
following effects:

| Action | Effect     |
|--------|------------|
| Hover mouse | Show x and y values |
| Drag mouse | Zoom x or y axis |
| Double click | Zoom out |

Several techniques from static graphics are also useful for
interactive graphics:

* **Multi-layer** graphics include 

# Examples

* [dygraphs](https://rstudio.github.io/dygraphs/gallery-synchronization.html)
  single-layer, multi-panel, single-plot, direct manipulation (click
  to zoom, hover to show data values and
  [tooltips](https://rstudio.github.io/dygraphs/gallery-annotations.html)).
  
# Summary of vocabulary

* **Actions**: direct manipuation (click, drag, hover), indirect
  manipulation (menus, buttons).
* **Effects**: zoom, highlight, show/hide (data, labels, tooltips).
* **Complexity**: multi-panel, multi-layer, multi-plot.

# Does not work yet

```{r, eval=FALSE}
library(highcharter)
data(diamonds, package = "ggplot2")
hchart(diamonds$price, color = "#B71C1C") %>%
  hc_title(text = "You can zoom me")
```
