# A vocabulary for characterizing interactive graphics

This tutorial presents several R packages for making interactive
graphics. To begin, please install the required packages:

```{r}
source("http://tdhock.github.io/interactive-tutorial/packages.R")
```

## Definition of interactive graphics

The R command line is interactive, so aren't all R graphics
interactive? Yes, if you have access to the data and R source
code. No, if you only have access to the graphic itself.

This tutorial uses the following definition: in an **interactive** or
**dynamic** graphic, the plot can be updated without changing the R
code. In a **non-interactive** or **static** graphic, the R code must
be changed to update the plot.

For example, the following is a static graphic

```{r}
lungDeaths <- cbind(mdeaths, fdeaths)
matplot(lungDeaths, type="l")
```

because it can be zoomed only by changing the R code:

```{r}
matplot(lungDeaths, type="l", xlim=c(20, 40))
```

In contrast, the following interactive graphic can be zoomed by
dragging the mouse on the plot (without changing the R code).

```{r}
dygraphs::dygraph(lungDeaths)
```

## Actions, effects, and complexity

We will use the terms **Actions** and **Effects** to categorize
different kinds of interactive graphics. In the graphic above, we see
several examples of **direct manipulation actions** which produce the
following effects:

| Action | Effect     |
|--------|------------|
| Hover mouse | Show x and y values |
| Drag mouse | Zoom x or y axis |
| Double click | Zoom out |

We use the term **direct manipulation** to describe **actions** near
the plotted data. This contrasts **indirect manipulation** on buttons
or menus which are typically located outside of the plotting region.

We will use some other terms to describe the **complexity** of a data
visualization. The complexity terms answer the following questions:

* How many layers? A single plot may have several layers or different
  geoms (points, lines, etc) that represent different parts of a
  complex data set. The data viz above is **single-layer** because it
  only displays one layer (lines).
* How many panels? A single plot may have several panels, each with
  its own axis. The data viz above is **single-panel** since it has
  only one panel.
* How many plots? A single data viz may have several linked plots,
  each with different axes and data. The data viz above is
  **single-plot** since it has only one plot.
  
## The shiny web server

The `shiny` package provides a web server that can be used for many
interactive graphics based on indirect manipulation actions. For
example, **multi-panel**, **multi-layer**, **single-plot** data viz
below uses **indirect manipulation** to show how the central limit
theorem works for data of different distributions.

```{r, eval=FALSE}
shiny::runGitHub('ShinySampleMean', 'ekstroem')
```

Another example with the same keywords to explain power calculations.

```{r, eval=FALSE}
shiny::runGitHub('ShinyPower', 'ekstroem')
```

## Multi-plot data visualization

This is not in R, but [dc.js](http://dc-js.github.io/dc.js/) can be
used to create **multi-plot**, **single-panel**, **single-layer** data
visualizations with **direct manipulation actions**.

## Animated data visualization

The `animation` and `animint` packages can be used to produce
**animated** data visualizations which are updated over time like a
video. The main difference is that

* `animation` uses **indirect manipulation**, and only allows
  scrolling forward or backward in time.
* `animint` uses **direct manipulation**, and allows interaction with
  any number of variables (not just the time variable).
  
Exercise: what keywords apply to the following data visualizations?

| Data | `animation`     | `animint` |
|--------|------------|--|
| World Bank | [viz+source](http://www.ggplot2-exts.org/gganimate.html) |[viz](http://cbio.ensmp.fr/~thocking/WorldBank-facets/) [source](https://github.com/tdhock/animint/blob/master/inst/examples/WorldBank-facets.R)| 
| Evolution | [viz](http://nicholsonppp.r-forge.r-project.org/2009-08-19/index.htm) |  [viz](http://cbio.ensmp.fr/~thocking/animint-evolution-plots/) [source](https://github.com/tdhock/animint/blob/master/inst/examples/evolution.R) | 

## Summary of vocabulary

We have discussed the following terms.

### Actions

* Direct manipulation (click, drag, hover) is useful for showing
  details of plotted data.
* Indirect manipulation (menus, buttons) is useful for changing the
  data subset or plotting parameters.
* Animation is useful for showing how data sets vary with a time
  variable.

### Effects

* Zooming is useful to show details of data that is already plotted.
* Showing/hiding data/labels/tooltips is useful to show details that
  are not already shown.
* Highlighting is useful to show the current selection. 

### Complexity

* Multi-layer graphics are useful when there are different data sets
  to compare on the same axes.
* Multi-panel graphics are useful in two situations.
  * Same geom for different data subsets (mostly useful for non-interactive graphics).
  * Different geoms for with aligned axes (useful for interactive graphics). 
* Multi-plot graphics are useful for data sets where there are more
  than two quantitative variables. You can show an overview in one
  plot and details in another.

