
# Breakpoint detection in copy number data

```{r}
data(intreg, package="animint")
str(intreg)
library(data.table)
(signals.dt <- data.table(intreg$signals))
```

## Multi-panel and multi-layer data viz

We can use facets to view each signal in a separate panel.

```{r}
gg.signals <- ggplot()+
  ggtitle("Copy number profiles")+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  scale_x_continuous(
    "position on chromosome (mega base pairs)",
    breaks=c(100, 200))+
  scale_y_continuous(
    "logratio (approximate DNA copy number)",
    breaks=c(-1, 0, 1))+
  geom_point(aes(base/1e6, logratio),
             shape=1,
             data=intreg$sig)
gg.signals+facet_wrap("signal")
gg.signals+facet_grid(signal ~ .)
gg.signals+facet_grid(. ~ signal)
gg.signals+facet_grid(. ~ signal, scales="free")
gg.signals+facet_grid(. ~ signal, scales="free", space="free")
```

Quiz: what keywords apply to the graphics above?

There are 20 segmentation models for each signal, which we can also
view in separate panels.

```{r}
gg.models <- gg.signals+
  geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                   showSelected=signal,
                   showSelected2=segments),
               data=intreg$seg,
               color="green")+
  geom_vline(aes(xintercept=base/1e6,
                 showSelected=signal,
                 showSelected2=segments),
             color="green",
             linetype="dashed",
             data=intreg$breaks)
gg.models+facet_grid(signal ~ segments, scales="free", space="free")
gg.models+facet_grid(segments ~ signal, scales="free", space="free")
```

Quiz: what keywords apply to the graphics above?

```{r}
library(data.table)
segments.dt <- data.table(intreg$segments)
setkey(segments.dt, signal, first.base, last.base)
signals.dt[, base.after := base+1L]
setkey(signals.dt, signal, base, base.after)
(signals.with.models <- foverlaps(signals.dt, segments.dt))
```

The data table above has joined segments with overlapping data points.

```{r}
model.selection <- signals.with.models[, list(
  total.squared.error=sum((logratio-mean)^2),
  data=.N), by=.(signal, segments)]
model.selection[, log.data := log(data)]
model.selection[, penalized.error := total.squared.error + log.data * segments]
model.selection[signal=="4.2",][order(segments),]
ggplot()+
  geom_line(aes(segments, total.squared.error, group=signal),
            data=model.selection)
ggplot()+
  geom_line(aes(segments, penalized.error, group=signal),
            data=model.selection)
addType <- function(dt, error.type){
  data.table(dt, error.type=factor(error.type, c("total squared error", "penalized error")))
}
ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "cm"))+
  facet_grid(error.type ~ ., scales="free")+
  ylab("")+
  geom_line(aes(segments, total.squared.error, group=signal),
            data=addType(model.selection, "total squared error"))+
  geom_line(aes(segments, penalized.error, group=signal),
            data=addType(model.selection, "penalized error"))
```

Quiz: what keywords apply to the graphics above?

Note that as the number of segments increases, the total squared error
always decreases, so attains its minimum at 20 segments. In contrast,
the penalized error (BIC) does not always decrease, so attains its
minimum at an intermediate number of segments. 

```{r}
sig.labels <- model.selection[segments==1,]
sig.seg.names <- signals
  group_by(signal, segments) %>%
  summarise(min.base=min(base),
            max.base=max(base)) %>%
  mutate(base=(min.base+max.base)/2)
sig.names <- sig.seg.names %>%
  group_by(signal, base) %>%
  summarise(logratio=max(signals.dt$logratio))
seg.names <- sig.seg.names %>%
  group_by(signal, segments, base) %>%
  summarise(logratio=min(signals.dt$logratio)-0.2)
tallrects <- make_tallrect(model.selection, "segments")
tallrects$geom_params$colour <- signal.colors[["estimate"]]
```

```{r}
library(animint)
data(intreg)
mmir.selection <- 
  list(segments=ggplot()+
       ggtitle("Select profile and number of segments")+
       tallrects+
       geom_text(aes(0, error, label=signal, clickSelects=signal),
                 data=sig.labels, hjust=1)+
       scale_x_continuous("segments", breaks=c(1, 5, 10, 20),
                          limits=c(-2, 20))+
       xlab("squared error")+
       geom_line(aes(segments, error,
                     group=signal,
                     clickSelects=signal),
                 data=model.selection,
                 alpha=0.6, size=8),

       signal=ggplot()+
       theme_animint(width=800)+       
       scale_x_continuous("position on chromosome (mega base pairs)",
                          breaks=c(100,200))+
       scale_fill_manual(values=breakpoint.colors,guide="none")+
       geom_blank(aes(first.base/1e6, logratio+2/8), data=intreg$ann)+
       ggtitle("Copy number profile and maximum likelihood segmentation")+
       ylab("logratio")+
       geom_point(aes(base/1e6, logratio,
                      showSelected=signal),
                  data=intreg$sig)+
       geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                        showSelected=signal,
                        showSelected2=segments),
                    data=intreg$seg, colour=signal.colors[["estimate"]])+
       geom_segment(aes(base/1e6, min(signals.dt$logratio),
                        xend=base/1e6, yend=max(signals.dt$logratio),
                        showSelected=signal,
                        showSelected2=segments),
                  colour=signal.colors[["estimate"]],
                  linetype="dashed",
                  data=intreg$breaks)+
       geom_text(aes(base/1e6, logratio, label=paste("signal", signal),
                     showSelected=signal),
                 data=sig.names)+
       geom_text(aes(base/1e6, logratio,
                     label=sprintf("%d segment%s", segments,
                       ifelse(segments==1, "", "s")),
                     showSelected=signal,
                     showSelected2=segments),
                 data=seg.names, color=signal.colors[["estimate"]]),

       first=list(signal="4.2", segments=4))
animint2dir(mmir.selection, "intreg-selection")
```

more examples.

```{r}
library(animint)

## Example: 4 plots, 2 selectors.
data(intreg)
signal.colors <- c(estimate="#0adb0a", latent="#0098ef")
breakpoint.colors <- c("1breakpoint"="#ff7d7d", "0breakpoints"='#f6f4bf')
model.linetypes <- c(margin="dotted",limit="dashed",regression="solid")
intreg$annotations$logratio <- max(intreg$sig$log)
## To get the bottom 3 plots to line up properly, we need to plot some
## geom_blanks bigger than the x range, so we calculate that here.
blank.items <- with(intreg,{
  list(segments=list(data=selection,x="min.L",y="segments"),
       error=list(data=selection,x="max.L",y="cost"),
       regression=list(data=model,x=c("min.L","max.L"),
                       y=c("min.feature","max.feature")),
       intervals=list(data=intervals,x=c("min.L","max.L"),y="feature"))
})
Lrange <- c()
for(N in names(blank.items)){
  L <- blank.items[[N]]
  Lrange <- range(c(Lrange,unlist(L$data[,L$x])),finite=TRUE)
  blank.items[[N]]$yrange <- range(unlist(L$data[,L$y]))
}
Lrange[1] <- Lrange[1]-1
Lrange[2] <- Lrange[2]+1
for(N in names(blank.items)){
  L <- blank.items[[N]]
  blank.items[[N]]$blank <- data.frame(x=Lrange, y=L$yrange)
}

mmir.plot <- 
  list(signal=ggplot()+
       theme_animint(height=300, width=800)+       
       scale_x_continuous("position on chromosome (mega base pairs)",
                          breaks=c(100,200))+
       geom_tallrect(aes(xmin=first.base/1e6, xmax=last.base/1e6,
                         fill=annotation,
                         showSelected=signal),
                     data=intreg$ann)+
       scale_fill_manual(values=breakpoint.colors,guide="none")+
       geom_text(aes((first.base+last.base)/2e6, logratio+1/8,
                     label=annotation,
                     showSelected=signal),
                 data=intreg$ann)+
       geom_blank(aes(first.base/1e6, logratio+2/8), data=intreg$ann)+
       geom_point(aes(base/1e6, logratio,
                      showSelected=signal),
                  data=intreg$sig)+
       geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                        showSelected=signal,
                        showSelected2=segments),
                    data=intreg$seg, colour=signal.colors[["estimate"]])+
       geom_vline(aes(xintercept=base/1e6,
                      showSelected=signal,
                      showSelected2=segments),
                  colour=signal.colors[["estimate"]],
                  linetype="dashed",
                  data=intreg$breaks),
       regression=ggplot()+
       theme_animint(height=150, width=800)+
       geom_blank(aes(x,y), data=blank.items$regression$blank)+
       geom_segment(aes(x=min.L, y=feature, xend=max.L, yend=feature,
                        clickSelects=signal),
                    size=5,
                    data=intreg$int)+
       geom_segment(aes(min.L, min.feature, xend=max.L, yend=max.feature,
                        linetype=line),
                    colour="red",
                    size=3,
                    data=intreg$model)+
       scale_linetype_manual(values=model.linetypes),
       error=ggplot()+
       theme_animint(height=100, width=800)+
       geom_blank(aes(x,y), data=blank.items$error$blank)+
       geom_segment(aes(min.L, cost, xend=max.L, yend=cost,
                        showSelected=signal), data=intreg$selection),
       segments=ggplot()+
       theme_animint(height=100, width=800)+
       geom_blank(aes(x,y), data=blank.items$segments$blank)+
       geom_segment(aes(min.L, segments, xend=max.L, yend=segments,
                        showSelected=signal), data=intreg$selection)+
       geom_tallrect(aes(xmin=min.L, xmax=max.L,
                         showSelected=signal,
                         clickSelects=segments),
                     data=intreg$selection,
                     alpha=1/2))
## This is a normal ggplot of all the data, subsets of which can be
## shown by clicking the plots.
sig.facets <- mmir.plot$sig+
  facet_grid(segments~signal, scales="free", space="free_x")+
  theme_bw()+
  theme(panel.margin=grid::unit(0,"cm"))
print(sig.facets)
animint2dir(mmir.plot, "intreg-nofacets")

## The mmir.plot above is way too complicated, since it does not use
## facets. The simpler facetted version looks like this:
mmir.facet <- 
  list(signal=mmir.plot$signal,
       
       penalty=ggplot()+
       geom_tallrect(aes(xmin=min.L, xmax=max.L,
                         showSelected=signal,
                         clickSelects=segments),
                     data=data.frame(intreg$selection, what="segments"),
                     alpha=1/2)+
       ylab("")+
       theme_bw()+
       theme_animint(height=500, width=800)+
       theme(panel.margin=grid::unit(0, "lines"))+
       geom_segment(aes(min.L, feature, xend=max.L, yend=feature,
                        clickSelects=signal),
                    size=5,
                    data=data.frame(intreg$int, what="regression"))+
       geom_segment(aes(min.L, min.feature, xend=max.L, yend=max.feature,
                        linetype=line),
                    colour="red",
                    size=3,
                    data=data.frame(intreg$model, what="regression"))+
       scale_linetype_manual(values=model.linetypes)+
       geom_segment(aes(min.L, cost, xend=max.L, yend=cost,
                        showSelected=signal),
                    data=data.frame(intreg$selection, what="error"))+
       geom_segment(aes(min.L, segments, xend=max.L, yend=segments,
                        showSelected=signal),
                    data=data.frame(intreg$selection, what="segments"))+
       xlab("penalty value $L=f(x)$")+ # TODO: mathjax.
       facet_grid(what~.,scales="free"),

       title="Max-margin interval regression")
animint2dir(mmir.facet, "intreg-facets") 
## This plot has an additional facet for signal, which would not be
## present in the interactive plot, but is useful here to see all
## the data in regular ggplot2.
too.many.facets <- mmir.facet$penalty+
  facet_grid(what~signal, scales="free")+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "cm"))
print(too.many.facets)

## Regions with linetype indicating errors.
breaks.by.signal <- split(intreg$breaks, intreg$breaks$signal)
anns.by.signal <- split(intreg$ann, intreg$ann$signal)
error.regions.list <- list()
for(signal in names(breaks.by.signal)){
  signal.breaks <- breaks.by.signal[[signal]]
  signal.anns <- anns.by.signal[[signal]]
  signal.anns$target.breaks <-
    ifelse(signal.anns$annotation=="1breakpoint", 1, 0)
  for(model.i in 1:20){
    model.breaks <- subset(signal.breaks, segments==model.i)
    signal.anns$breaks <- NA
    for(region.i in 1:nrow(signal.anns)){
      region <- signal.anns[region.i, ]
      after.start <- region$first.base < model.breaks$base
      before.end <- model.breaks$base < region$last.base
      signal.anns$breaks[region.i] <- sum(after.start & before.end)
    }
    signal.anns$error.type <- with(signal.anns, {
      ifelse(breaks < target.breaks, "false negative",
             ifelse(target.breaks < breaks, "false positive", "correct"))
    })
    error.regions.list[[paste(model.i, signal)]] <-
      data.frame(segments=model.i, signal.anns)
  }
}
error.regions <- do.call(rbind, error.regions.list)

reg <- subset(intreg$model, line=="regression")
slope <- with(reg, (min.L-max.L)/(min.feature-max.feature))
intreg$intervals$pred.L <-
  slope * (intreg$intervals$feature - reg$min.feature) + reg$min.L
intreg.errors <- 
  list(signal=ggplot()+
       theme_animint(height=300, width=800)+       
       scale_x_continuous("position on chromosome (mega base pairs)",
                          breaks=c(100,200))+
       ylab("noisy copy number logratio signal")+
       geom_tallrect(aes(xmin=first.base/1e6, xmax=last.base/1e6,
                         fill=annotation,
                         linetype=error.type,
                         showSelected2=segments,
                         showSelected=signal),
                     color="black",
                     alpha=0.5,
                     data=error.regions)+
       scale_linetype_manual("error type", values=c(correct=0,
                                   "false negative"=3,
                                   "false positive"=1))+
       guides(linetype=guide_legend(override.aes=list(fill="white")))+       
       scale_fill_manual(values=breakpoint.colors)+
       geom_blank(aes(first.base/1e6, logratio+2/8), data=intreg$ann)+
       geom_point(aes(base/1e6, logratio,
                      showSelected=signal),
                  data=intreg$sig)+
       geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                        showSelected=signal,
                        showSelected2=segments),
                    data=intreg$seg, colour=signal.colors[["estimate"]])+
       geom_vline(aes(xintercept=base/1e6,
                      showSelected=signal,
                      showSelected2=segments),
                  colour=signal.colors[["estimate"]],
                  linetype="dashed",
                  data=intreg$breaks),
       penalty=ggplot()+
         theme_bw()+
         theme_animint(height=500, width=800)+
         theme(panel.margin=grid::unit(0, "cm"))+
       geom_tallrect(aes(xmin=min.L, xmax=max.L,
                         showSelected=signal,
                         clickSelects=segments),
                     data=data.frame(intreg$selection
                      ##,what="segments"
                                     ),
                     alpha=1/2)+
       ylab("")+
       geom_vline(aes(xintercept=pred.L, showSelected=signal),
                  color="violet",
                  data=intreg$intervals)+
       geom_segment(aes(min.L, feature, xend=max.L, yend=feature,
                        clickSelects=signal),
                    size=6,
                    data=data.frame(intreg$intervals, what="regression"))+
       geom_segment(aes(min.L, min.feature, xend=max.L, yend=max.feature,
                        linetype=line),
                    colour="violet",
                    size=3,
                    data=data.frame(intreg$model, what="regression"))+
       ## geom_point(aes(pred.L, feature),
       ##            size=5,
       ##            data=data.frame(intreg$intervals, what="regression"))+
       scale_linetype_manual(values=model.linetypes)+
       geom_segment(aes(min.L, cost, xend=max.L, yend=cost,
                        showSelected=signal),
                    data=data.frame(intreg$selection, what="incorrect labels"))+
       geom_segment(aes(min.L, segments, xend=max.L, yend=segments,
                        showSelected=signal),
                    data=data.frame(intreg$selection, what="segments"))+
       xlab("penalty value log(lambda)")+ # TODO: mathjax.
       facet_grid(what~., scales="free"),
       title="Max-margin interval regression")
animint2dir(intreg.errors, "intreg-errors")

##animint2gist(intreg.errors)

## No annotated regions!
mmir.segs <- 
  list(signal=ggplot()+
       theme_animint(height=300, width=800)+       
       scale_x_continuous("position on chromosome (mega base pairs)",
                          breaks=c(100,200))+
       scale_fill_manual(values=breakpoint.colors,guide="none")+
       geom_blank(aes(first.base/1e6, logratio+2/8), data=intreg$ann)+
       ggtitle("Copy number profile and maximum likelihood segmentation")+
       ylab("logratio")+
       geom_point(aes(base/1e6, logratio,
                      showSelected=signal),
                  data=intreg$sig)+
       geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                        showSelected=signal,
                        showSelected2=segments),
                    data=intreg$seg, colour=signal.colors[["estimate"]])+
       geom_vline(aes(xintercept=base/1e6,
                      showSelected=signal,
                      showSelected2=segments),
                  colour=signal.colors[["estimate"]],
                  linetype="dashed",
                  data=intreg$breaks),
       
       segments=ggplot()+
       theme_animint(height=300, width=800)+
       xlab("log(penalty)")+
       ylab("optimal number of segments")+
       ggtitle("Select profile and number of segments")+
       geom_tallrect(aes(xmin=min.L, xmax=max.L,
                         showSelected=signal,
                         clickSelects=segments),
                     data=intreg$selection,
                     alpha=1/2)+
       geom_segment(aes(min.L, segments, xend=max.L, yend=segments,
                        clickSelects=signal),
                    data=intreg$selection, alpha=0.6, size=5))
animint2dir(mmir.segs, "intreg-segs")

library(reshape2)
model.tall <- melt(model.selection, measure.vars=c("error", "penalized.error"))
## Plot error AND penalized error versus number of segments.
mmir.buggy <- 
  list(segments=ggplot()+
       ggtitle("Select profile and number of segments")+
       tallrects+
       scale_x_continuous("segments", breaks=c(1, 5, 10, 20),
                          limits=c(-2, 21))+
       xlab("")+
       facet_grid(variable ~ ., scales="free_y")+
       geom_line(aes(segments, value,
                     group=interaction(signal, variable),
                     clickSelects=signal),
                 data=model.tall,
                 alpha=0.6, size=8),

       signal=ggplot()+
       theme_animint(width=800)+       
       scale_x_continuous("position on chromosome (mega base pairs)",
                          breaks=c(100,200))+
       scale_fill_manual(values=breakpoint.colors,guide="none")+
       geom_blank(aes(first.base/1e6, logratio+2/8), data=intreg$ann)+
       ggtitle("Copy number profile and maximum likelihood segmentation")+
       ylab("logratio")+
       geom_point(aes(base/1e6, logratio,
                      showSelected=signal),
                  data=intreg$sig)+
       geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                        showSelected=signal,
                        showSelected2=segments),
                    data=intreg$seg, colour=signal.colors[["estimate"]])+
       geom_segment(aes(base/1e6, min(signals.dt$logratio),
                        xend=base/1e6, yend=max(signals.dt$logratio),
                        showSelected=signal,
                        showSelected2=segments),
                  colour=signal.colors[["estimate"]],
                  linetype="dashed",
                  data=intreg$breaks)+
       geom_text(aes(base/1e6, logratio, label=paste("signal", signal),
                     showSelected=signal),
                 data=sig.names)+
       geom_text(aes(base/1e6, logratio,
                     label=sprintf("%d segment%s", segments,
                       ifelse(segments==1, "", "s")),
                     showSelected=signal,
                     showSelected2=segments),
                 data=seg.names, color=signal.colors[["estimate"]]),

       first=list(signal="4.2", segments=4))
animint2dir(mmir.buggy, "intreg-buggy")

## Plot error AND penalized error versus number of segments.
penalized <- model.selection %>%
  group_by(signal) %>%
  filter(penalized.error < penalized.error[1]*2)
mmir.BIC <- 
  list(segments=ggplot()+
       ggtitle("Select profile and number of segments")+
       tallrects+
       scale_x_continuous("segments", breaks=c(1, 5, 10, 20),
                          limits=c(-2, 21))+
       xlab("")+
       facet_grid(variable ~ ., scales="free_y")+
       geom_line(aes(segments, error,
                     group=signal,
                     clickSelects=signal),
                 data=data.frame(model.selection,
                   variable="un-penalized error"),
                 alpha=0.6, size=8)+
       geom_line(aes(segments, penalized.error,
                     group=signal,
                     clickSelects=signal),
                 data=data.frame(penalized, variable="penalized error (BIC)"),
                 alpha=0.6, size=8),

       signal=ggplot()+
       theme_animint(width=800)+       
       scale_x_continuous("position on chromosome (mega base pairs)",
                          breaks=c(100,200))+
       scale_fill_manual(values=breakpoint.colors,guide="none")+
       geom_blank(aes(first.base/1e6, logratio+2/8), data=intreg$ann)+
       ggtitle("Copy number profile and maximum likelihood segmentation")+
       ylab("logratio")+
       geom_point(aes(base/1e6, logratio,
                      showSelected=signal),
                  data=intreg$sig)+
       geom_segment(aes(first.base/1e6, mean, xend=last.base/1e6, yend=mean,
                        showSelected=signal,
                        showSelected2=segments),
                    data=intreg$seg, colour=signal.colors[["estimate"]])+
       geom_segment(aes(base/1e6, min(signals.dt$logratio),
                        xend=base/1e6, yend=max(signals.dt$logratio),
                        showSelected=signal,
                        showSelected2=segments),
                  colour=signal.colors[["estimate"]],
                  linetype="dashed",
                  data=intreg$breaks)+
       geom_text(aes(base/1e6, logratio, label=paste("signal", signal),
                     showSelected=signal),
                 data=sig.names)+
       geom_text(aes(base/1e6, logratio,
                     label=sprintf("%d segment%s", segments,
                       ifelse(segments==1, "", "s")),
                     showSelected=signal,
                     showSelected2=segments),
                 data=seg.names, color=signal.colors[["estimate"]]),

       first=list(signal="4.2", segments=4))
animint2dir(mmir.BIC, "intreg-BIC")
```
